%\VignetteEngine{knitr::docco_classic} 
%\VignetteIndexEntry{quickstart} 

```{r include=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE, message=FALSE, error=FALSE, dev='svg', fig.show='hold')
```

Quickstart with rivr
====================

Overview
--------

`rivr` provides functions for calculating the normal depth and critical depth,
computing gradually-varied flow profiles, and solving the unsteady flow 
equations using finite-differencing methods. Outputs are formatted for 
manipulation with `dplyr` and visualization with `ggplot2`. This vignette 
provides a brief introduction to the main functions and example applications.

Class methods
-------------

The `rivr` package defines a new class `rivr`, which is essentially an 
extension of the `data.frame` class with additional atributes. The following
methods are defined for class `rivr`:

  - `head` 
  - `tail`
  - `print`
  - `summary`
  - `plot`

   

Computing normal and critical depth
-----------------------------------

The normal depth $y_n$ is defined as the flow depth at which
$$
Q = \frac{C_m}{n} AR^{2/3}S_0^{1/2}
$$
Where $Q$ is the flow rate, $n$ is Manning's coefficient, $A$ is the 
cross-sectional flow area (also a function of flow depth), $R$ is the 
hydraulic radius and $S_0$ is the bed slope. The critical depth $y_c$ is 
defined as the flow depth at which 
$$
\frac{dE}{dy} = 1 - \frac{Q^2}{gA^3}\frac{dA}{dy} = 0.
$$
Both $y_n$ and $y_c$ are non-linear functions of $y$. `rivr` provides 
functions for computing normal and critical depths, as shown below. In both 
cases a Newton-Raphson scheme is used to solved the equations.

```{r ynyc}
require(rivr)
flow = 250; mannings = 0.045 ; Cm = 1.486; gravity = 32.2
width = 100; slope = 0.001; sideslope = 0

yn = normal_depth(slope, mannings, flow, yopt = 2, Cm, width, sideslope)
yc = critical_depth(flow, yopt = 2, gravity, width, sideslope)

print(c(normal.depth = yn, critical.depth = yc))
```

Standard-step method for gradually-varied flow profiles
-------------------------------------------------------

The standard step method can be used to solve steady-state water surface 
profiles. The solution to gradually-varied flow profiles is based on the 
non-linear ordinary differential equation
$$
\frac{dy}{dx} = \frac{S_0 - S_f}{1 - Fr^2}
$$
and is appropriate for cases where $\frac{dy}{dx}$ is small. The standard-step
method operates by stepping along the channel by a constant distance interval,
starting from a cross-section where the flow depth is known (the control section). 
The flow depth is computed at the adjacent cross-section (target section). The 
computed value at the target is then used as the basis for computing flow depth at 
the next cross-section, i.e. the previous target section becomes the new control 
section for each step. A Newton-Raphson scheme is used each step to compute the 
flow depth and friction slope. Technically, the average friction slope of the 
control and target section is used to compute the flow depth at the target section. 

```{r gvf}
flow = 250; mannings = 0.045 ; Cm = 1.486; gravity = 32.2
width = 100; slope = 0.001; sideslope = 0

gvf = compute_profile(slope, mannings, flow, y0 = 2.7, Cm, gravity, width, 
  sideslope, stepdist=50, totaldist=3000)
```

While a `plot` method is defined for `rivr` objects, the output is also 
formatted for easy visualization with `ggplot2`.

```{r plot-gvf}
require(ggplot2)
ggplot(gvf, aes(x = x, y = y + z)) + geom_line(color='blue')
# or try the default plot method
# plot(gvf)
```

Unsteady flow
-------------

Unsteady flow models solve the shallow water equations (conservation of mass 
and conservation of momentum). Kinematic wave models (KWM) use a truncated form 
of the momentum equation while dynamic wave models (DWM) solve the mass and
momentum equations simultaneously. A variety of numerical schemes can be used to 
solve these equations. `rivr` provides an interface to multiple explicit finite 
differencing schemes (one KWM scheme and two DWM schemes) for computing unsteady
flows. The MacCormack scheme is recommended for most applications. 

```{r uf}
baseflow = 250; mannings = 0.045 ; Cm = 1.486; gravity = 32.2
width = 100; slope = 0.001; sideslope = 0

numnodes = 301; xresolution = 250; tresolution = 10; 
times = seq(0, 30000, by = tresolution)
wave = ifelse(times >= 9000, baseflow,
  baseflow + (750/pi)*(1 - cos(pi*times/(60*75))))
downstream = rep(-1, length(wave))

mn = c(1, 101, 201)
mt = c(501, 1501, 3001)
uf = route_wave(slope, mannings, Cm, gravity, width, sideslope, 
  baseflow, wave, downstream, tresolution, xresolution, numnodes, 
  mn, mt, "Dynamic", "MacCormack", "QQ") 
```

Methods are defined for printing and summarizing `rivr` objects. Accessing the 
unsteady flow outputs for in-depth analysis is greatly simplified by using 
`dplyr`.

```{r uf-access}
require(dplyr)
uf.nodes = filter(uf, monitor.type == "node")
ggplot(uf.nodes, aes(x=time, y=flow, color=factor(distance))) + geom_line()
uf.times = filter(uf, monitor.type == "timestep")
ggplot(uf.times, aes(x=distance, y=flow, color=factor(time))) + geom_line()
# or try the default plot method
# plot(uf)
```
